Synthesizing Quotient Lenses

### 2018-07-02 papers/QuotientLenses/main.tex
Converted to camera ready state

\documentclass[acmsmall,review,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For double-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
% % For single-blind review submission, w/o CCS and ACM Reference (max
% submission space)
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
% % For single-blind review submission, w/ CCS and ACM Reference
% \documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true} % For
% final camera-ready submission, w/ required CCS and ACM Reference
% \documentclass[acmsmall]{acmart}\settopmatter{}

% % Journal information % Supplied to authors by publisher for camera-ready
% submission; % use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{ICFP} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

% % Copyright information % Supplied to authors (based on authors' rights
% management selection; % see authors.acm.org) by publisher for camera-ready
% submission; % use 'none' for review submission.
\setcopyright{none}
\usepackage{algorithm, amsmath, amssymb, verbatim, enumerate, graphicx,
  centernot, tikz, array, mathtools, bussproofs, stmaryrd, enumitem,
  stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\usepackage{proof}
\usepackage{xspace}
\usepackage[noend]{algpseudocode}
\usepackage[capitalize]{cleveref}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\keywords{quotient regular expressions, QRE lens, synthesis}
\maketitle

\section{Introduction}
Programmers often need to write programs that bidirectionally convert data
between two different formats. For example, bibliographic data may be stored in
\bibtex{} or EndNote formats, spreadsheet data can be represented as either CSV
and TSV files, and web APIs can return data as either JSON or XML objects.  A
time-tested and appealing way to implement such conversions is to use
\emph{lenses}~\cite{Lenses}, programs that simultaneously define pairs of
functions for translating the data from the source format to the target format
(the \emph{get} direction) and back again (the \emph{put} direction).  Lens
programming languages guarantee that these \kw{put} and \kw{get} functions
satisfy certain \emph{lens laws}, which imply that round trips, from source to
target and back again, behave properly.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks
from the benchmark suite of the original Optician system. These
benchmarks are a combination of custom benchmarks, benchmarks derived
from FlashFill~\cite{flashfill}, and benchmarks derived from
Augeas~\cite{augeas2} (we also experimented using our QRE implementation to synthesize quotient lenses between XML, RDF and and JSON formats using data from data.gov database; the data consisted of census statistics, demographic statistics, wage comparosion data, and crime index data).  In these 39 benchmarks, 10 of
the data formats had to be modified to work with the bijectivity
constraints that Optician required. For instance, when one
representation permits whitespace where the other does not, we
modified the original version of the benchmark to allow more whitespace,
thereby restoring bijectivity (but altering the data format). With the
new QRE support, we were able to remove these alterations.  This
experience alone suggests that QREs make the lens development process
more flexible.

@@

\bibliographystyle{plain}
\bibliography{local}

\appendix
\addcontentsline{toc}{section}{Appendices}
\section*{Appendices}
\section{Proof of Normal Form Theorem for QRE lenses}
\cref{normal form} claims that if there is a derivation $q : Q_1
\Leftrightarrow Q_2$, then there exists a bijective lens $\ell : K(Q_1)
\Leftrightarrow K(Q_2)$ such that:
\begin{align*}
\llbracket q \rrbracket.\get &= \llbracket \ell \rrbracket\circ
\canonize(Q_1)\\
\llbracket q \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1} \circ
\canonize(Q_2)
\end{align*}
We now prove this theorem.
\begin{proof}
We proceed by induction over the derivation $q : Q_1 \Leftrightarrow Q_2$.
\begin{enumerate}
  \item
  $\kw{lift}(\ell): R/\mathit{id}(R) \Leftrightarrow S/\mathit{id}(S)$ where
  $\ell : R \Leftrightarrow S$. Then:
  \begin{align*}
\llbracket \kw{lift}(\ell) \rrbracket.\get &=  \llbracket \ell \rrbracket
= \llbracket \ell \rrbracket \circ id_{\mathcal{L}(R)} =
\llbracket \ell \rrbracket \circ \canonize(\mathit{id}(R)), \text{ and }\\
\llbracket \kw{lift}(\ell) \rrbracket.\lput &= \llbracket \ell
\rrbracket^{-1} = \llbracket \ell \rrbracket^{-1} \circ id_{\mathcal{L}(S)} =
\llbracket \ell \rrbracket^{-1} \circ \canonize(id(S))
\end{align*}
\item
$\lquot(Q_1, q): Q_1 \; ; \; Q_2 \Leftrightarrow Q_3$ where $q : Q_2
\Leftrightarrow Q_3$, $Q_1$ is well formed and $K(Q_1) = W(Q_2)$. Then:
\begin{align*}
\llbracket \lquot(Q_1, q) \rrbracket.\get  &= \llbracket q
\rrbracket.\get \circ \canonize(Q_1)\\
\llbracket \lquot(Q_1, q) \rrbracket.\lput &= \llbracket q \rrbracket.\lput
\end{align*}
By the induction hypothesis, there exists a bijective lens $\ell :
K(Q_2) \Leftrightarrow K(Q_3)$ such that:
\begin{align*}
\llbracket q \rrbracket.\get &= \llbracket \ell \rrbracket \circ
\canonize(Q_2)\\
\llbracket q \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1} \circ
\canonize(Q_3)
\end{align*}
Consequently
\begin{align*}
\llbracket \lquot(Q_1, q)\rrbracket.\get  &= (\llbracket \ell \rrbracket \circ
\canonize(Q_2)) \circ \canonize(Q_1) = \llbracket \ell \rrbracket \circ
(\canonize(Q_1 \; ; \; Q_2))\\
\llbracket \lquot(Q_1, q) \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1}
\circ \canonize(Q_3)
\end{align*}

@@

\llbracket q_1 \; ; \; q_2 \rrbracket.\get &= \llbracket q_2
\rrbracket.\get\circ \llbracket q_1 \rrbracket.\get, \text{ and }\\
\llbracket q_1 \; ; \; q_2 \rrbracket.\lput &= \llbracket q_1 \rrbracket.\lput

@@

\llbracket q_1 \rrbracket.\get &= \llbracket \ell_1 \rrbracket \circ
\llbracket q_1 \rrbracket.\lput &= {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_1)
and
\llbracket q_2 \rrbracket.\get &= \llbracket \ell_2 \rrbracket \circ

@@

\llbracket q_2 \rrbracket.\lput &= {\llbracket \ell_2 \rrbracket}^{-1} \circ

@@

@@

A similar argument shows that:
$$\llbracket q_1 \; ; \; q_2 \rrbracket.\lput =

@@

\canonize(c)$$

@@

\llbracket q^* \rrbracket.\get &= (\llbracket q \rrbracket.\get)^*, \text{
and }\\
\llbracket q^* \rrbracket.\lput &= (\llbracket q \rrbracket.\lput)^*

@@

\llbracket q \rrbracket.\get &= \llbracket \ell \rrbracket \circ
\llbracket q \rrbracket.\lput &= {\llbracket \ell \rrbracket}^{-1} \circ

@@

\llbracket q_1 \cdot q_2 \rrbracket.\get &= \llbracket q_1 \rrbracket.\get \cdot
\llbracket q_2 \rrbracket.\get, \text{ and }\\
\llbracket q_1 \cdot q_2 \rrbracket.\lput &= \llbracket q_1 \rrbracket.\lput

@@

\llbracket q \rrbracket.\get &= \llbracket \ell_1 \rrbracket \circ
\llbracket q \rrbracket.\lput &= {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_3)
and
\llbracket q_2 \rrbracket.\get &= \llbracket \ell_2 \rrbracket \circ

@@

\llbracket q_2 \rrbracket.\lput &= {\llbracket \ell_2 \rrbracket}^{-1} \circ

@@

Similarly:

@@

\rrbracket^{-1} \circ \canonize(Q_3 \cdot Q_4) $$

@@

\llbracket q_1 \rrbracket.\get &= \llbracket \ell_1 \rrbracket \circ
\llbracket q_1 \rrbracket.\lput &= {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_3)
and
\llbracket q_2 \rrbracket.\get &= \llbracket \ell_2 \rrbracket \circ

@@

\llbracket q_2 \rrbracket.\lput &= {\llbracket \ell_2 \rrbracket}^{-1} \circ

@@

\rrbracket)^{-1} \circ (\canonize(Q_4) \cdot \canonize(Q_3))$$

@@

\llbracket q_1 \rrbracket.\get &= \llbracket \ell_1 \rrbracket \circ
\llbracket q_1 \rrbracket.\lput &= {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_3)
and
\llbracket q_2 \rrbracket.\get &= \llbracket \ell_2 \rrbracket \circ

@@

\ell_2 \rrbracket \circ \canonize(Q_1 \sep Q_2)$. A similar argument shows
that $\llbracket q_1 \sep q_2 \rrbracket.\lput = \llbracket \ell_1 \sep

@@

>>>>

\documentclass[acmsmall,screen]{acmart}

\pdfpagewidth=8.5in 
\pdfpageheight=11in

\settopmatter{}

\setcopyright{none}
\acmPrice{}
\acmDOI{10.1145/3234603}
\acmYear{2018}
\copyrightyear{2018}
\acmJournal{PACMPL}
\acmVolume{2}
\acmNumber{ICFP}
\acmArticle{14}
\acmMonth{9}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011050.10011017</concept_id>
<concept_desc>Software and its engineering~Domain specific languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011066.10011070</concept_id>
<concept_desc>Software and its engineering~Application specific development environments</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Domain specific languages}
\ccsdesc[300]{Software and its engineering~Application specific development environments}

\keywords{Bidirectional Programming, Program Synthesis, Type-Directed Synthesis,
Type Systems}

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}

\author{Solomon Maina}
\affiliation{
  \institution{University of Pennsylvnia}
  \country{USA}
}
\email{smaina@seas.upenn.edu}

\author{Anders Miltner}
\affiliation{
  \institution{Princeton University}
  \country{USA}
}
\email{amiltner@cs.princeton.edu}

\author{Kathleen Fisher}
\affiliation{
  \institution{Tufts University}
  \country{USA}
}
\email{kfisher@eecs.tufts.edu}

\author{Benjamin C. Pierce}
\affiliation{
  \institution{University of Pennsylvania}
  \country{USA}
}
\email{bcpierce@cis.upenn.edu}

\author{David Walker}
\affiliation{
  \institution{Princeton University}
  \country{USA}
}
\email{dpw@cs.princeton.edu}

\author{Steve Zdancewic}
\affiliation{
  \institution{University of Pennsylvania}
  \country{USA}
}
\email{stevez@cis.upenn.edu}

\usepackage{algorithm, amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz, array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption,flushend,relsize, listings, proof, xspace, booktabs}
\usepackage[noend]{algpseudocode}
\usepackage[capitalize]{cleveref}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}

\captionsetup{compatibility=false}

@@

\clubpenalty = 10000
\widowpenalty = 10000
\displaywidowpenalty = 10000
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\maketitle
\renewcommand{\shortauthors}{S. Maina, A. Miltner, K. Fisher, B. C. Pierce, D. Walker, S. Zdancewic}
\section{Introduction}
Programmers often need to write programs that bidirectionally convert data
between two different formats. For example, bibliographic data may be stored in
\bibtex{} or EndNote formats, spreadsheet data can be represented as either CSV
and TSV files, and web APIs can return data as either JSON or XML objects.  A
time-tested and appealing way to implement such conversions is to use
\emph{lenses}~\cite{Lenses}, programs that simultaneously define pairs of
functions for translating the data from the source format to the target format
(the \emph{get} direction) and back again (the \emph{put} direction).  Lens
programming languages guarantee that these \kw{put} and \kw{get} functions
satisfy certain \emph{lens laws}, which imply that round trips, from source to
target and back again, behave properly.

@@

To evaluate our QRE implementation, we adapted 39 lens synthesis tasks
from the benchmark suite of the original Optician system. These
benchmarks are a combination of custom benchmarks, benchmarks derived
from FlashFill~\cite{flashfill}, and benchmarks derived from
Augeas~\cite{augeas2} (we also experimented using our QRE implementation to synthesize quotient lenses between XML, RDF and and JSON formats using data from the data.gov database; the data consisted of census statistics, demographic statistics, wage comparosion data, and crime index data).  In these 39 benchmarks, 10 of
the data formats had to be modified to work with the bijectivity
constraints that Optician required. For instance, when one
representation permits whitespace where the other does not, we
modified the original version of the benchmark to allow more whitespace,
thereby restoring bijectivity (but altering the data format). With the
new QRE support, we were able to remove these alterations.  This
experience alone suggests that QREs make the lens development process
more flexible.

@@

\appendix
\addcontentsline{toc}{section}{Appendices}
\section*{Appendices}
\section{Proof of Normal Form Theorem for QRE lenses}
\cref{normal form} claims that if there is a derivation $q : Q_1
\Leftrightarrow Q_2$, then there exists a bijective lens $\ell : K(Q_1)
\Leftrightarrow K(Q_2)$ such that:
$$
\llbracket q \rrbracket.\get = \llbracket \ell \rrbracket\circ
\canonize(Q_1) \text{ and } 
\llbracket q \rrbracket.\lput = \llbracket \ell \rrbracket^{-1} \circ
\canonize(Q_2)
$$
We now prove this theorem.
\begin{proof}
We proceed by induction over the derivation $q : Q_1 \Leftrightarrow Q_2$.
\begin{enumerate}
  \item
  $\kw{lift}(\ell): R/\mathit{id}(R) \Leftrightarrow S/\mathit{id}(S)$ where
  $\ell : R \Leftrightarrow S$. Then:
  \begin{align*}
\llbracket \kw{lift}(\ell) \rrbracket.\get &=  \llbracket \ell \rrbracket
= \llbracket \ell \rrbracket \circ id_{\mathcal{L}(R)} =
\llbracket \ell \rrbracket \circ \canonize(\mathit{id}(R)), \text{ and }\\
\llbracket \kw{lift}(\ell) \rrbracket.\lput &= \llbracket \ell
\rrbracket^{-1} = \llbracket \ell \rrbracket^{-1} \circ id_{\mathcal{L}(S)} =
\llbracket \ell \rrbracket^{-1} \circ \canonize(id(S))
\end{align*}
\item
$\lquot(Q_1, q): Q_1 \; ; \; Q_2 \Leftrightarrow Q_3$ where $q : Q_2
\Leftrightarrow Q_3$, $Q_1$ is well formed and $K(Q_1) = W(Q_2)$. Then:
$$
\llbracket \lquot(Q_1, q) \rrbracket.\get  = \llbracket q
\rrbracket.\get \circ \canonize(Q_1) \text{ and }
\llbracket \lquot(Q_1, q) \rrbracket.\lput = \llbracket q \rrbracket.\lput
$$
By the induction hypothesis, there exists a bijective lens $\ell :
K(Q_2) \Leftrightarrow K(Q_3)$ such that:
$$
\llbracket q \rrbracket.\get = \llbracket \ell \rrbracket \circ
\canonize(Q_2) \text{ and }
\llbracket q \rrbracket.\lput = \llbracket \ell \rrbracket^{-1} \circ
\canonize(Q_3)
$$
Consequently
\begin{align*}
\llbracket \lquot(Q_1, q)\rrbracket.\get  &= (\llbracket \ell \rrbracket \circ
\canonize(Q_2)) \circ \canonize(Q_1) = \llbracket \ell \rrbracket \circ
(\canonize(Q_1 \; ; \; Q_2))\\
\llbracket \lquot(Q_1, q) \rrbracket.\lput &= \llbracket \ell \rrbracket^{-1}
\circ \canonize(Q_3)
\end{align*}

@@

\llbracket q_1 \; ; \; q_2 \rrbracket.\get = \llbracket q_2
\rrbracket.\get\circ \llbracket q_1 \rrbracket.\get \text{ and }
\llbracket q_1 \; ; \; q_2 \rrbracket.\lput = \llbracket q_1 \rrbracket.\lput

@@

\[\begin{array}{l l}
\llbracket q_1 \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(c) & \llbracket q_2 \rrbracket.\get = \llbracket \ell_2 \rrbracket \circ

@@

\llbracket q_1 \rrbracket.\lput = {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_1) & \llbracket q_2 \rrbracket.\lput = {\llbracket \ell_2 \rrbracket}^{-1} \circ

@@

\end{array}
\]

@@

A similar argument shows that $\llbracket q_1 \; ; \; q_2 \rrbracket.\lput =

@@

\canonize(c)$

@@

\llbracket q^* \rrbracket.\get = (\llbracket q \rrbracket.\get)^* \text{ and }
\llbracket q^* \rrbracket.\lput = (\llbracket q \rrbracket.\lput)^*

@@

\llbracket q \rrbracket.\get = \llbracket \ell \rrbracket \circ
\canonize(Q_1) \text{ and }
\llbracket q \rrbracket.\lput = {\llbracket \ell \rrbracket}^{-1} \circ

@@

\llbracket q_1 \cdot q_2 \rrbracket.\get = \llbracket q_1 \rrbracket.\get \cdot
\llbracket q_2 \rrbracket.\get \text{ and }
\llbracket q_1 \cdot q_2 \rrbracket.\lput = \llbracket q_1 \rrbracket.\lput

@@

\[\begin{array}{l l}
\llbracket q \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(Q_1) & \llbracket q_2 \rrbracket.\get = \llbracket \ell_2 \rrbracket \circ

@@

\llbracket q \rrbracket.\lput = {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_3) & \llbracket q_2 \rrbracket.\lput = {\llbracket \ell_2 \rrbracket}^{-1} \circ

@@

Similarly, $

@@

\rrbracket^{-1} \circ \canonize(Q_3 \cdot Q_4)$

@@

\[\begin{array}{l l}
\llbracket q_1 \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(Q_1) & \llbracket q_2 \rrbracket.\get = \llbracket \ell_2 \rrbracket \circ

@@

\llbracket q_1 \rrbracket.\lput = {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonize(Q_3) & \llbracket q_2 \rrbracket.\lput = {\llbracket \ell_2 \rrbracket}^{-1} \circ

@@

\rrbracket)^{-1} \circ (\canonize(Q_4) \cdot \canonize(Q_3))$

@@

\[\begin{array}{l l }
\llbracket q_1 \rrbracket.\get = \llbracket \ell_1 \rrbracket \circ
\canonize(Q_1) & \llbracket q_2 \rrbracket.\get = \llbracket \ell_2 \rrbracket \circ

@@

\ell_2 \rrbracket \circ \canonize(Q_1 \sep Q_2)$. Similarly, $\llbracket q_1 \sep q_2 \rrbracket.\lput = \llbracket \ell_1 \sep

@@

\bibliography{local}
